---
title: "NoSQL"
output:
  html_document:
    df_print: paged
---



# Rational DBs vs Nosql Databases

The most common type of nosql databases is document database.

- table vs collection
- row vs document

# Types of NoSQL DBs.

- **Key-value stores** are the simplest. Every item in the database is stored as an attribute name (or "key") together with its value. Riak, Voldemort, and Redis are the most well-known in this category.
- **Wide-column stores** store data together as columns instead of rows and are optimized for queries over large datasets. The most popular are Cassandra and HBase.
- **Document databases** pair each key with a complex data structure known as a document. Documents can contain many different key-value pairs, or key-array pairs, or even nested documents. MongoDB is the most popular of these databases.
- **Graph databases** are used to store information about networks, such as social connections. Examples are Neo4J and HyperGraphDB.

Some references:

- https://www.mongodb.com/scale/types-of-nosql-databases
- https://neo4j.com/developer/graph-db-vs-nosql/


# Airbnb sample data

MongoDB is the most popular NOSQL database.

https://docs.atlas.mongodb.com/sample-data/sample-airbnb/

To connect to this MongoDB, you need to either on the campus network or connect via UCDavis VPN.

```{r}
library(tidyverse)
library(mongolite)

m <- mongo("airbnb", db = "data", url = "mongodb://mongouser:secret@alan.ucdavis.edu/data")
```

```{r}
# total number of documents
m$count()
```


```{r}
# total number of documents satisfying a condition
m$count('{"price": {"$lt": 20}}')
```


Query Documents
https://docs.mongodb.com/manual/tutorial/query-documents/


```{r}
# find the first document
first_doc <- m$find("{}", limit = 1)
first_doc %>% glimpse()
```

```{r}
# filter fields
m$find("{}", fields = '{"name": true}')
```

```{r}
# filter keys
m$find('{"price": {"$lt": 20}}', fields = '{"name": true, "price": true}')
```

Check for all the operators: https://docs.mongodb.com/manual/reference/operator/query-comparison/#query-selectors-comparison

```{r}
m$find('{"price": {"$lt": 50}, "bedrooms": {"$gte": 3}, "amenities": "Wifi"}',
  fields = '{"name": true, "price": true, "bedrooms": true}'
)
```

Remark: As `amenities` is an array, `"amenities": "Wifi"` means Wifi is one of the amenities. On the other hand, `"amenities": ["Wifi"]` means Wifi is the only amenity.

```{r}
# sort
m$find('{"bedrooms": {"$gte": 4}}',
  fields = '{"name": true, "price": true}',
  sort = '{"price": -1}',
  limit = 5
)
```


## Query nested fields

```{r}
countrydf <- m$find(
  '{"address.country_code": "US"}',
  fields = '{"name": true, "address.country": true}'
)
countrydf
```

```{r}
countrydf$address$country
```

```{r}
countrydf %>% pull(address) %>% pull(country)
```

```{r}
countrydf %>%
  mutate(country = address$country, address = NULL)
```
```{r}
countrydf %>% 
  bind_cols(countrydf$address) %>% 
  select(-address)
```


## Query an array

```{r}
# find all documents where amenities contain both Oven and Refrigerator
m$find('{"amenities": {"$all": ["Oven", "Refrigerator"]}}',
       fields = '{"name": true, "amenities": true}')
```

PS: you could also use the `$and` operator.

```{r}
# find all documents where amenities contain both Oven and Refrigerator
m$find('{"$and": [{"amenities": "Oven"}, {"amenities": "Refrigerator"}]}',
       fields = '{"name": true, "amenities": true}')
```


```{r}
# find all documents where amenities contain Oven or Refrigerator or both
m$find('{"amenities": {"$in": ["Oven", "Refrigerator"]}}',
       fields = '{"name": true, "amenities": true}')
```
PS: you could also use the `$or` operator.

```{r}
# find all documents where amenities contain both Oven and Refrigerator
m$find('{"$or": [{"amenities": "Oven"}, {"amenities": "Refrigerator"}]}',
       fields = '{"name": true, "amenities": true}')
```

## Regular expressions (come back to this later)

```{r}
m$find('{"name": {"$regex": "\\\\bsea\\\\b", "$options": "i"}}', fields = '{"name": true}')
```

## Iteration

```{r}
m$find('{"price": {"$lt": 12}}')
```

```{r}
it <- m$iterate('{"price": {"$lt": 12}}')
while (TRUE) {
  x <- it$one()
  if (is.null(x)) break
  cat(x$name, "\n")
}
```

By batch

```{r}
it <- m$iterate('{"price": {"$lt": 12}}')
res <- it$batch(size = 2)
```

Examine raw JSON

```{r}
it <- m$iterate('{"price": {"$lt": 12}}')
```

```{r}
j <- it$json(size = 1)
```


## Aggregate

The aggregation pipeline is a framework for data aggregation modeled on the concept of data processing pipelines.

https://docs.mongodb.com/manual/reference/operator/aggregation-pipeline/

### Group

To illustrate the usage of a pipeline, consider the following three stages

- filter the data
- group by and summerize

```{r}
m$aggregate('[
  {"$match": {"bedrooms": {"$ne": null, "$ne": 0}}},
  {"$group": { "_id": "$bedrooms", "avg_price": { "$avg": "$price" }}}
]')
```

User the `$project` stage to rename `_id`.
```{r}
m$aggregate('[
  {"$match": {"bedrooms": {"$ne": null, "$ne": 0}}},
  {"$group": { "_id": "$bedrooms", "avg_price": { "$avg": "$price" }}},
  {"$project": { "_id": false, "bedrooms": "$_id", "avg_price": true }},
  {"$sort": {"bedrooms": 1}}
]')
```

Group by two variables
```{r}
m$aggregate('[
  {"$match": {"bedrooms": {"$ne": null, "$ne": 0}}},
  {"$group": {
    "_id": {"property_type": "$property_type", "bedrooms": "$bedrooms"},
    "avg_price": { "$avg": "$price" }}},
  {"$project": {
    "_id": false,
    "property_type": "$_id.property_type",
    "bedrooms": "$_id.bedrooms",
    "avg_price": true
  }},
  {"$sort": {"property_type": 1, "bedrooms": 1}}
]')
```

### Aggregation Pipeline Operators

Compute the average price per room grouped by property_type

See: https://docs.mongodb.com/manual/reference/operator/aggregation/

Remark: our mongo server is 3.x.

```{r}
m$aggregate('[
  {"$match": {"bedrooms": {"$ne": null, "$ne": 0}}},
  {"$project": {
    "property_type": "$property_type",
    "bedrooms": "$bedrooms",
    "price_per_room": {"$divide": ["$price", "$bedrooms"]}
  }},
  {"$group": {
    "_id": "$property_type",
    "avg_bedrooms": { "$avg": "$bedrooms" },
    "price_per_room": { "$avg": "$price_per_room" }}
  },
  {"$project": {
    "_id": false,
    "property_type": "$_id",
    "avg_bedrooms": true,
    "price_per_room": true
  }},
  {"$sort": {"property_type": 1}}
]')
```

### Unwind an array

`$unwind` has a similar effect as `tidyverse`'s `unnest_longer`.

```{r}
df <- tibble(
  id = c("a", "b"),
  v = list(1:3, 5:6),
  w = c("apple", "banana")
)
```
```{r}
df %>% unnest_longer(v)
```


To query all the reviews of "Ribeira Charming Duplex".
```{r}
m$aggregate('[
  {"$match": {"name": "Ribeira Charming Duplex"}},
  {"$project": {
    "name": true,
    "reviews": true
  }},
  {"$unwind": "$reviews"},
  {"$project": {
    "_id": false,
    "name": true,
    "reviewer_name": "$reviews.reviewer_name",
    "comments": "$reviews.comments"
  }}
]')
```

### Iteration

```{r}
it <- m$aggregate('[
  {"$match": {"bedrooms": {"$ne": null, "$ne": 0}}},
  {"$group": { "_id": "$bedrooms", "avg_price": { "$avg": "$price" }}},
  {"$project": { "_id": false, "bedrooms": "$_id", "avg_price": true }},
  {"$sort": {"bedrooms": 1}}
]', iterate = TRUE)
```

```{r}
it$batch(size = 2)
```

