---
title: "Data Wrangling"
output: 
  html_document: 
    toc: yes
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
```

Quote from the book "R for Data Science", the author said

> R is an old
language, and some things that were useful 10 or 20 years ago now get in your
way. It’s difficult to change base R without breaking existing code, so most
innovation occurs in packages.

## What is tidyverse?

- base R's functions are often slow and the implementations are often not consistent
- writing code in `tidyverse` style usually is more elegant
- the operations would be easily chained together using piping (more below)

So what is tidyverse?

- It is a collection of R packages which are designed to be used together. 
  - `ggplot2`, for data visualisation
  - `dplyr`, for data manipulation
  - `tidyr`, for data tidying
  - `readr`, for data import
  - `purrr`, for functional programming
  - `tibble`, for tibbles, a modern re-imagining of data frames
  - `stringr`, for strings
  - `forcats`, for factors


## Pipe operator

We are going to detour a bit to talk about pipe operator.
Pipe operator is a powerful way to make your code much cleaner and readible by

- structuring sequences of data operations left-to-right (as opposed to from the inside and out),
- avoiding nested function calls,
- minimizing the need for local variables and function definitions, and
- making it easy to add steps anywhere in the sequence of operations.

The operators pipe their left-hand side values forward into expressions that
appear on the right-hand side, i.e. one can replace `f(x)` with 
`x %>% f()`, where `%>%` is the (main) pipe-operator. When coupling 
several function calls with the pipe-operator, the benefit will become
more apparent. Consider this pseudo example:

```{r, eval = FALSE}
raw_data <- read.csv("/path/to/data/file.csv")
sub_data <- subset(raw_data, variable_a > x)
new_data <- transform(sub_data, variable_c = variable_a / variable_b)
the_data <- head(new_data, 100)
```
However, the local variables are really not necessary, so one liner could be

```{r, eval = FALSE}
the_data <- head(
  transform(
    subset(
      read.csv("/path/to/data/file.csv"),
      variable_a > x
    ),
    variable_c = variable_a / variable_b
  ),
  100
)
```

But that is too hard to read and reason. If we use pipe operator,

```{r, eval = FALSE}
the_data <- read.csv("/path/to/data/file.csv") %>%
  subset(variable_a > x) %>%
  transform(variable_c = variable_a / variable_b) %>%
  head(100)
```

Hint: In RStudio, you could use `Ctrl + Shift + M` (or `Cmd + Shift + M`) to insert the pipe operator.


### Basic usage

* `x %>% f` or `x %>% f()` is equivalent to `f(x)`
* `x %>% f(y)` is equivalent to `f(x, y)`
* `x %>% f %>% g %>% h` is equivalent to `h(g(f(x)))`

```{r}
c(1, 2) %>% sum()
```

```{r}
choose(5, 3)
5 %>% choose(3)
```


### The argument placeholder

* `x %>% f(y, .)` is equivalent to `f(y, x)`
* `x %>% f(y, z = .)` is equivalent to `f(y, z = x)`

```{r}
3 %>% choose(5, .)
3 %>% choose(5, k = .)
```

### Re-using the placeholder

It is straightforward to use the placeholder several times
in a right-hand side expression. However, when the placeholder
only appears in a nested expressions magrittr will still apply
the first-argument rule. The reason is that in most cases this
results more clean code. 

`x %>% f(y = nrow(.), z = ncol(.))` is equivalent to 
   `f(x, y = nrow(x), z = ncol(x))`

```{r}
5 %>% choose(. - 2)
```

The behavior can be
overruled by enclosing the right-hand side in braces:

`x %>% {f(y = nrow(.), z = ncol(.))}` is equivalent to 
   `f(y = nrow(x), z = ncol(x))`

```{r}
list(n = 5, k = 3) %>% {
  choose(.$n, .$k)
}
```


## `dplyr` basics

- It offers five basic verbs
  - `select`: picks variables based on their names
  - `filter`: picks cases based on their values
  - `mutate`: adds new variables that are functions of existing variables
  - `arrange`: changes the ordering of the rows
  - `summarize` or `summarise`: reduces multiple values down to a single summary

- These all combine naturally with `group_by` which allows you to perform any operation "by group". 


## Obtain some data

First of all, we need some data to work with. If the data is stored in a `csv`,

```{r, eval = FALSE}
flights <- read_csv("flights.csv")
```

We are using the `tidyverse` function `read_csv` to import the `flights.csv` instead of the obsolete base function `read.csv`.
- `read_csv` imports data as `tibble` which has better and more consistent handling of variables.
- `read_csv` is often faster than `read.csv`
- `read_csv` handles unicode characters better

There are also `read_tsv` and `read_delim` for reading tab-seperated or delimited files.


The datasets are actually obtained from the R package `nycflights13`

```{r}
# Airline on-time data for all flights departing NYC in 2013.
library(nycflights13)
flights
```

## `select`: picks variables based on their names.

To select arrival and departure times,
```{r}
# old way to do it
# flights[, c("arr_time", "dep_time")]
flights %>% select(arr_time, dep_time)
```

### I don't see why it's useful

`dplyr` provides a lot of helper functions,

```{r}
# colon `:` specifies all the variables between the columns of `dep_time` and `arr_time`
flights %>% select(dep_time:arr_time)
# all the columns start with arr_
flights %>% select(starts_with("arr_"))
# all the columns end with _time
flights %>% select(ends_with("_time"))
# all the columns contain dep
flights %>% select(contains("dep"))
# remove particular columns
flights %>% select(-year)
# all the columns do not contain dep
flights %>% select(-contains("dep"))
# using regular expression (later of the course)
flights %>% select(matches("^(arr|dep)_"))
```

Related verbs:
- if you just need a single variable, you could use `pull`.
- you could use `rename` to rename columns

```{r}
# flights$arr_time
flights %>% pull(arr_time)
```

```{r}
flights %>% rename(y = year)
```


```{r}
# of course, we could select everything
flights %>% select(everything())
# move air_time to the front
flights %>% select(air_time, everything())
```


## `filter`: picks cases based on their values

```{r}
# flights[, flights$origin == "JFK"]
# subset(flights, origin == "JFK")
flights %>% filter(origin == "JFK")
flights %>% filter(distance > 1000)
# note that we are using a single `&` instead of `&&` as in base R
flights %>% filter(origin == "JFK" & distance > 1000)
flights %>% filter(origin == "JFK", distance > 1000)
flights %>% filter(distance < 500 | distance > 1000)
flights %>% filter(!between(distance, 500, 1000))
# only keep the complete cases
flights %>% filter(complete.cases(.)) # filter(flights, complete.cases(flights))
flights %>% drop_na() # same, but more efficient
```

## Chaining and piping

Very often, we will need to use multiple `dplyr` verbs, for example 
```{r}
# the pipe operator %>% increases readability
flights %>%
  select(origin, air_time) %>%
  filter(origin == "JFK") %>%
  filter(air_time < 500) %>%
  rename(airtime = air_time)
```
```{r}
# slightly more efficient
flights %>%
  select(origin, airtime = air_time) %>%
  filter(origin == "JFK", airtime < 500)
```



```{r}
# a few more examples
flights %>%
  select(origin, air_time) %>%
  filter(origin == "JFK", air_time < mean(air_time, na.rm = TRUE))

# maybe we want to calculate mean_air_time first
mean_air_time <- flights %>%
  pull(air_time) %>%
  mean(na.rm = TRUE)
flights %>%
  select(origin, air_time) %>%
  filter(origin == "JFK", air_time > mean_air_time)
```

```{r}
# what if there is a name collision?
air_time <- flights %>%
  pull(air_time) %>%
  mean(na.rm = TRUE)
flights %>%
  select(origin, air_time) %>%
  filter(origin == "JFK", air_time > !!air_time)
```

Remarks: use `slice` if you want particular rows

```{r}
flights %>% slice(100:105)
```

## `mutate`: adds new variables that are functions of existing variables

```{r}
flights %>% mutate(
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
# we could refer to the columns just created
flights %>% mutate(
  gain = dep_delay - arr_delay,
  gain_per_hour = gain / (air_time / 60)
)
```
```{r}
# remove a variable
flights %>% mutate(
  gain = dep_delay - arr_delay,
  gain_per_hour = gain / (air_time / 60),
  air_time = NULL
)
```

```{r}
# `transmute` only keep the new variables
flights %>% transmute(
  gain = dep_delay - arr_delay,
  gain_per_hour = gain / (air_time / 60)
)
```

### Six variations on ranking functions

- `row_number`: equivalent to `rank(ties.method = "first")`
- `min_rank`: equivalent to `rank(ties.method = "min")`
- `dense_rank`: like `min_rank()`, but with no gaps between ranks
- `percent_rank`: a number between 0 and 1 computed by rescaling min_rank to [0, 1]
- `cume_dist`: a cumulative distribution function. Proportion of all values less than or equal to the current rank.
- `ntile`: a rough rank, which breaks the input vector into n buckets

```{r}
some_data <- tibble(
  x = c(3, 4, 1, 3, 1)
)
some_data %>%
  mutate(
    row_number(),
    row_number(x),
    min_rank(x),
    dense_rank(x),
    percent_rank(x),
    cume_dist(x),
    ntile(x, 3)
  )
```

### `lead` and `lag`

```{r}
some_data2 <- tibble(
  time = 1:5,
  value = c(3, 4, 1, 3, 1)
)
some_data2 %>% mutate(
  diff1 = value - lag(value), 
  diff2 = lead(value) - value
)
```



### Conditional mutation

```{r}
flights %>% transmute(
  arr_delay = arr_delay,
  status = if_else(arr_delay > 0, "delayed", "on time")
)
(flight_distances <- flights %>%
  transmute(
    distance,
    distance_type = case_when(
      distance < 500 ~ "short",
      distance < 1000 ~ "mid",
      TRUE ~ "long"
    )
  ))
```

### `recode` values

```{r}
flight_distances %>% mutate(distance_type = recode(distance_type,
  "long" = "long-distance",
  "mid" = "mid-distance",
  "short" = "short-distance"
))
```

```{r}
some_data %>% mutate(
  y = recode(x,
    `1` = 10,
    `3` = 30
  )
)
```


## `arrange`: changes the ordering of the rows

```{r}
flights %>% arrange(year, month, day)
flights %>% arrange(desc(dep_delay))
```

## `summarize` and `group_by` operations

```{r}
flights %>%
  group_by(tailnum) %>%
  summarize(n = n())
flights %>%
  group_by(tailnum) %>%
  tally() # shorthand
flights %>% count(tailnum) # another shorthand
```

```{r}
flights %>%
  group_by(tailnum) %>%
  summarize(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  )
```


```{r}
flights %>%
  group_by(dest) %>%
  summarize(
    planes = n_distinct(tailnum),
    flights = n()
  )
```


```{r}
# group by multiple variables
(per_day <- flights %>%
  group_by(year, month, day) %>%
  summarize(flights = n()))
```

```{r}
(per_month <- per_day %>%
  summarize(flights = sum(flights)))
```

```{r}
(per_year <- per_month %>%
  summarize(flights = sum(flights)))
```

### `across` columns

The `across` function requires dplyr 1.0, which at the point of writing, has not been released yet.

Interested reader could install the development version of `dplyr` with
```r
# take your own risk to install it!!
devtools::install_github("tidyverse/dplyr")
```
A toolchain would be required to install it from source.

See https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-colwise/


Suppose, we want to compuate the means for the all numeric columns of `flights`. Naively, we could do

```{r}
flights %>%
  group_by(origin) %>%
  summarize(
    dep_time = mean(dep_time, na.rm = TRUE),
    sched_dep_time = mean(sched_dep_time, na.rm = TRUE),
    dep_delay = mean(dep_delay, na.rm = TRUE),
    arr_time = mean(arr_time, na.rm = TRUE),
    sched_arr_time = mean(sched_arr_time, na.rm = TRUE),
    arr_delay = mean(arr_delay, na.rm = TRUE),
    flight = mean(flight, na.rm = TRUE),
    air_time = mean(air_time, na.rm = TRUE),
    distance = mean(distance, na.rm = TRUE)
  )
```

With the new `across` function, it is as easy as:
```{r}
flights %>%
  group_by(origin) %>%
  summarize(across(dep_time:distance, mean, na.rm = TRUE))
```

We have accidentally included some non numeric columns.

```{r}
flights %>%
  group_by(origin) %>%
  summarize(
    across(dep_time:distance & is.numeric, mean, na.rm = TRUE)
  )
```

We could also use predicate functions in `across` and use different functions to summerize different variables.
```{r}
flights %>%
  group_by(origin) %>%
  summarize(
    across(ends_with("time"), mean, na.rm = TRUE),
    across(ends_with("delay"), mean, na.rm = TRUE, trim = 0.2),
  )
```

## Other useful functions


```{r}
flights %>% glimpse()
flights %>% 
  sample_n(1000) %>% 
  ggplot(aes(arr_delay, dep_delay)) + geom_point(na.rm = TRUE)
## rows with largest values of air_time with the original order preserved
flights %>% top_n(3, air_time)
# two disadvantages: order changes and does not handle ties
flights %>% arrange(desc(air_time)) %>% head(3)
```


## Two-table verbs

There are 6 types of joins.

- `inner_join(x, y)` only includes observations that match in both x and y
- `left_join(x, y)` includes all observations in x, regardless of whether they match or not. 
- `right_join(x, y)` equivalent to left_join(y, x)
- `full_join(x, y)` includes all observations from x and y
- `semi_join(x, y)` only keep observations from x if there is a match in y
- `anti_join(x, y)` remove observations from x if there is a match in y

```{r}
(df1 <- tibble(id = c(1, 2), v = 2:1))
(df2 <- tibble(id = c(1, 3), a = 10, b = "a"))
```

```{r}
df1 %>% inner_join(df2, by = "id")
```


```{r}
df1 %>% left_join(df2, by = "id")
```


```{r}
df1 %>% right_join(df2, by = "id")
```


```{r}
# outer join
df1 %>% full_join(df2, by = "id")
```


```{r}
df1 %>% semi_join(df2, by = "id")
```


```{r}
df1 %>% anti_join(df2, by = "id")
```

### Join by the same variable of different names

```{r}
df1 <- tibble(id = c(1, 2), v = 2:1)
df2 <- tibble(name = c(1, 3), a = 10, b = "a")
```

```{r}
df1 %>% inner_join(df2, by = c("id" = "name"))
```


## Row bind or column bind data frames

```{r}
df1 <- tibble(a = "hello", b = 2, c = 3)
df2 <- tibble(a = "world", b = 4)
```

```{r}
df1 %>% bind_rows(df2)
```


The `bind_rows` function from `dplyr` is better than `rbind` in several ways.

- `bind_rows` handles missing values
- `bind_rows` doesn't automatically convert types


```{r, error = TRUE}
df1 <- tibble(a = "hello", b = 2, c = 3)
df2 <- tibble(a = "world", b = 4)
df1 %>% rbind(df2)
```


```{r, error = TRUE}
df1 <- tibble(a = "hello", b = 2)
df2 <- tibble(a = 2, b = 4)
df1 %>% rbind(df2)
df1 %>% bind_rows(df2)
```

Likewise, `bind_cols` is safer than `cbind`.

```{r}
df1 <- tibble(x = 1:5)
df2 <- tibble(y = 1:10)
```

```{r}
df1 %>% cbind(df2)
```

```{r, error = TRUE}
df1 %>% bind_cols(df2)
```


## Crossing and nesting

Crossing allows you to examine all the combinations between variables.

For example, we know that not all routes between `origin` and `dest` exist.
```{r}
all_existing_routes <- flights %>% expand(nesting(origin, dest))
all_possible_routes <- flights %>% expand(crossing(origin, dest))
# or simply  
all_possible_routes <- flights %>% expand(origin, dest)
```

```{r}
# these routes do not exist
all_possible_routes %>% 
  anti_join(all_existing_routes)
```

## Completion and missing values

`expand` only keeps on the variables specified. If you want to join the original data, you could use `complete`.

```{r}
dat <- flights %>% 
  group_by(origin, dest) %>% 
  summarize(air_time = mean(air_time, na.rm = TRUE))
```


```{r}
dat %>%
  ungroup() %>%  # dplyr 0.8.5
  complete(origin, dest, fill = list("air_time" = Inf))
```

```{r}
df <- tribble(
  ~x, ~y,
   1,  2,
  NA,  3,
   3, NA
)
df
```

```{r}
# remove all rows with missing values
df %>% drop_na()
```


```{r}
# fill values by copying downward
df %>% fill(x, y)
```


```{r}
df %>% 
  replace_na(list(x = Inf, y = -Inf))
```


## Pivoting Data

Using the datasets from R for Data Science to show that the same data could be organized in different ways.

In general, there are two formats,

- wide format
  - each subject has its own row
  - a variable may take multiple columns
- long format
  - each variable has its own column
  - a subject may take multiple rows

### Pivot longer

```{r}
relig_income
```

```{r}
# make sure you have tidyr 1.0
relig_income %>%
  pivot_longer(-religion, names_to = "income", values_to = "count")
```

```{r}
billboard
```

```{r}
billboard %>%
  pivot_longer(
    starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

### Pivot wider

```{r}
fish_encounters
```

```{r}
fish_encounters %>% pivot_wider(
  names_from = station,
  values_from = seen,
  values_fill = list(seen = 0)
)
```

```{r}
us_rent_income
```

```{r}
us_rent_income %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe))
```


## References

- R for Data Science http://r4ds.had.co.nz/tidy-data.html
- Documentation of dplyr https://dplyr.tidyverse.org/
- Documentation of tidyr https://tidyr.tidyverse.org/
