---
title: "Web Scraping"
output: 
  html_document: 
    toc: yes
---


```{r}
library(tidyverse)
library(rvest)
```


# Web scraping

Web scraping is a technique of extracting useful information from webpages. But why do you need it? 
Not all companies provide an API, some times, you will need to scrape information directly from their websites.


# HTML and XML

Here is an example of a simple HTML page:

```html
<!DOCTYPE html>
<html>
<head>
<title>Page Title</title>
</head>
<body>
<h1>This is a Heading</h1>
<p>This is a paragraph.</p>
</body>
</html>
```

Of course, there are a lot of different tags. Go to https://www.w3schools.com/html/ to see some html basics.
Nevertheless, all these tags are predefined so your browser knows what each tag means.

(We only need to only a bit HTML in order to do web scrapping)


## Look at web page source code

It is important to identify the thing that you want to scrape in the webpage. The best way to do it is to use the inspect function in the Chrome browser.


## CSS selector

CSS selector is the most common way to locate an object in a website. 

- `tag`: the name of the tag, for example `td` and `a` tags
- `.class`: class of an object
- `#id`: id of an object


## imdb example

Suppose we want to get the list of most top rated movies from https://www.imdb.com/chart/top/?ref_=nv_mv_250

We see that all movies names are under the `<td>/<a>` nodes. The `<td>` nodes have class name `titleColumn`.

```{r}
html <- read_html("https://www.imdb.com/chart/top/?ref_=nv_mv_250")
```


```{r}
# it finds all the <td> nodes, but we only need the node with class `titleColumn`
td_nodes <- html %>% html_nodes("td")
# it finds all the <td> nodes with class titleColumn
title_columns <- html %>% html_nodes("td.titleColumn")
# it finds all the <a> nodes within <td> nodes with class titleColumn
a_nodes <- title_columns %>% html_node("a")
# html_text to get the values inside the <a> </a> tag
movie_names <- a_nodes %>% html_text()
head(movie_names)
```

Put everything together
```{r}
movie_names <- html %>%
  html_nodes("td.titleColumn") %>% 
  html_node("a") %>%
  html_text()
```


Now, we also want to capture the ratings.

```{r}
imdb_ratings <-  html %>%
  html_nodes("td.ratingColumn.imdbRating") %>%
  html_node("strong") %>%
  html_text()
```

```{r}
tibble(title = movie_names, rating = imdb_ratings)
```

How if you also want to get the years? There is also a cute function `html_table`.

```{r}
html %>% html_node("table.chart.full-width") %>%
  html_table() %>% 
  as_tibble(.name_repair = "unique") %>%
  select(rank_and_title = `Rank & Title`, rating = `IMDb Rating`) %>%
  separate(rank_and_title, c("rank", "title", "year"), sep = "\n")
```



Now, we want to url link to the movie "The Shawshank Redemption".

```{r}
shawshank_url <- html %>%
  html_nodes("td.titleColumn") %>%
  html_node("a") %>%
  keep(html_text(.) == "The Shawshank Redemption") %>%
  html_attr("href")
```

Alternatively, we could the CSS pseudo class `contains()`

```{r}
shawshank_url <- html %>%
  html_node('td.titleColumn a:contains("The Shawshank Redemption")') %>%
  html_attr("href")
```

But it is not the complete url, we need to base url.
```{r}
shawshank_full_url <- paste0("https://www.imdb.com/", shawshank_url)
```

Then we could further scrape things from `shawshank_full_url`.



## Stackoverflow example

Besides using node class, you could also search a node by its `id`.

Here we are first extracting the `div` node with `id="questions"`.

```{r}
read_html("https://stackoverflow.com/questions/tagged/r") %>%
  html_node("div#questions") %>%
  html_nodes("div.summary") %>%
  html_nodes("h3") %>%
  html_nodes("a") %>%
  html_text()
```


# Scraping dynamic webpages

`rvest` is only able to scrape static web pages. If you want to scrape dynamic web pages, you will need to control a browser programatically.

PS: Please note that dynamic scraping is not suitable to be used in shiny application.


## UC Davis class search


```{r}
library(RSelenium)
requireNamespace("wdman")
requireNamespace("httpuv")
```


```{r}
port <- httpuv::randomPort()
# find a version of chromedriver that is compatible with your chrome
# binman::list_versions("chromedriver")
server <- wdman::chrome(port = port, version = "87.0.4280.87", verbose = FALSE)
rd <- remoteDriver(port = port)
```


```{r}
rd$open(silent = TRUE)
# check if there is any error messages
stopifnot(is.null(rd$sessionInfo$message))

rd$navigate("https://registrar-apps.ucdavis.edu/courses/search/index.cfm")
```

```{r}
rd$findElement("css", "#course_number")$clickElement()
rd$sendKeysToActiveElement(list("STA"))
rd$findElement(
    "css", 
    '#home_tablez_bz input[name="search"]'
  )$clickElement()
```


```{r}
retry::retry({
  rd$getPageSource()[[1]] %>% 
    read_html() %>% 
    html_node("div#courseResultsDiv table") %>% 
    html_table() %>% {
      df <- slice(., -(1:4))
      names(df) <- slice(., 4)
      df
    } %>% 
    separate("CRN Time/Days", c("CRN", "Time/Days"), "\n\\s+\t")
  },
  when = "xml_missing",
  timeout = 5
)
```

```{r}
# close the browser finally
server$stop()
```

## NBA site

PS: actually nba.com has a undocumented API, see https://github.com/seemethere/nba_py/wiki/stats.nba.com-Endpoint-Documentation


```{r}
port <- httpuv::randomPort()
server <- wdman::chrome(port = port, version = "87.0.4280.87", verbose = FALSE)
rd <- remoteDriver(port = port)
```

```{r}
rd$open(silent = TRUE)
rd$navigate("https://stats.nba.com/leaders/?SeasonType=Regular%20Season")
```

```{r}
retry::retry({
  rd$getPageSource()[[1]] %>%
    read_html() %>%
    html_node("div.nba-stat-table__overflow table") %>%
    html_table()
  },
  when = "xml_missing",
  timeout = 5
)  
```

loop over the table by clicking the next button
```{r}
leader <- NULL
for (i in 1:6) {
  leader <- bind_rows(
    leader,
    rd$getPageSource()[[1]] %>%
      read_html() %>%
      html_node("div.nba-stat-table__overflow table") %>%
      html_table()
  )
  nextbutton <- rd$findElement("css", "a.stats-table-pagination__next")
  nextbutton$clickElement()
}
```

```{r}
leader
```

```{r}
# close the browser finally
server$stop()
```

## Advanced topics

- Headless browser
- XPaths

